% ============================================================
% 03b_joint_estimation_fisher_matrix.tex
% ============================================================

\section{Joint Estimation: Fisher Information Matrix for $(t, D, \sigma_0)$}
\label{sec:joint_estimation}

A common criticism of single-parameter bounds is their reliance on unrealistically
strong assumptions about known nuisance parameters. In this section we remove this
limitation by deriving the full Fisher information matrix (FIM) for joint estimation
of time $t$, diffusion coefficient $D$, and observation noise floor $\sigma_0$.

\subsection{Observation Model}

We consider the Gaussian observation model introduced previously,
\begin{equation}
p(x \mid t, D, \sigma_0)
=
\mathcal{N}\!\left(0,\, s(t,D,\sigma_0)\, I_d \right),
\qquad
s(t,D,\sigma_0) := 2Dt + \sigma_0^2 .
\end{equation}

The parameter vector is
\[
\theta := (t, D, \sigma_0^2),
\]
where $\sigma_0^2$ is used instead of $\sigma_0$ to ensure smoothness and to avoid
spurious singularities in the Fisher information.

\subsection{Score Functions}

The log-likelihood for a single observation is
\[
\log p(x \mid \theta)
=
-\frac{d}{2}\log(2\pi s)
-
\frac{\|x\|^2}{2s}.
\]

The score functions are
\begin{align}
\partial_t \log p
&=
D\left(
\frac{\|x\|^2}{s^2}
-
\frac{d}{s}
\right),
\\[6pt]
\partial_D \log p
&=
t\left(
\frac{\|x\|^2}{s^2}
-
\frac{d}{s}
\right),
\\[6pt]
\partial_{\sigma_0^2} \log p
&=
\frac{1}{2}
\left(
\frac{\|x\|^2}{s^2}
-
\frac{d}{s}
\right).
\end{align}

All three score functions are proportional to the same sufficient statistic,
reflecting an intrinsic degeneracy of the inference problem.

\subsection{Fisher Information Matrix}

The Fisher information matrix for one observation is
\[
J_{ij} := \mathbb{E}\!\left[
(\partial_{\theta_i}\log p)
(\partial_{\theta_j}\log p)
\right].
\]

Using Gaussian moment identities, we obtain
\begin{equation}
J^{(1)}
=
\frac{2d}{s^2}
\begin{pmatrix}
D^2 & Dt & \frac{D}{2} \\[6pt]
Dt & t^2 & \frac{t}{2} \\[6pt]
\frac{D}{2} & \frac{t}{2} & \frac{1}{4}
\end{pmatrix}.
\end{equation}

This matrix has rank one, reflecting the fact that all parameters enter the model
only through the single scalar combination $s=2Dt+\sigma_0^2$.

For $N_{\mathrm{eff}}$ independent or effectively independent observations,
\[
J = N_{\mathrm{eff}} J^{(1)}.
\]

\subsection{Cramér--Rao Bound for Time Under Joint Estimation}

Although the full parameter vector is not jointly identifiable from a single
snapshot, meaningful marginal bounds can still be obtained. The covariance matrix
of any unbiased estimator satisfies
\[
\mathrm{Cov}(\hat\theta) \succeq J^{-1},
\]
where the inverse is understood in the sense of the Moore--Penrose pseudoinverse.

The marginal lower bound for the variance of $\hat t$ is therefore
\[
\mathrm{Var}(\hat t)
\ge
(J^{-1})_{tt}.
\]

Direct inversion yields
\begin{equation}
(J^{-1})_{tt}
=
\frac{s^2}{2d N_{\mathrm{eff}} D^2}
\left(
1 + \frac{\sigma_0^2}{2Dt}
\right).
\end{equation}

Equivalently,
\begin{equation}
\mathrm{Var}(\hat t)
\ge
\frac{(2Dt + \sigma_0^2)^2}{2d D^2 N_{\mathrm{eff}}}
\left(
1 + \frac{\sigma_0^2}{2Dt}
\right).
\end{equation}

\subsection{Interpretation}

The additional factor
\[
1 + \frac{\sigma_0^2}{2Dt}
\]
quantifies the degradation of temporal precision due to joint estimation of nuisance
parameters. In the diffusion-dominated regime ($2Dt \gg \sigma_0^2$), this factor
approaches unity and the single-parameter CRLB is recovered. In the PSF-dominated
regime, the penalty diverges, reflecting the fundamental non-identifiability of $t$
when spatial broadening is dominated by observation noise.

\subsection{Identifiability and Degeneracy}

The structure of the Fisher information matrix makes explicit that time $t$, diffusion
coefficient $D$, and noise floor $\sigma_0^2$ cannot be independently identified from
a single snapshot without additional constraints, calibration, or multi-time data.
This degeneracy is not a weakness of the bound but an intrinsic property of the
inference problem and reinforces the operational nature of time distinguishability.
