% ============================================================
% 05_photon_limited_regime.tex
% ============================================================

\section{Photon-Limited Regime and Self-Consistent Temporal Resolution}
\label{sec:photon_limited}

In many experimental settings—single-particle tracking (SPT), fluorescence correlation
spectroscopy (FCS), and localization microscopy—the dominant limitation is not the number
of particles but the finite photon flux. In this regime, the effective sample size
$N_{\mathrm{eff}}$ is itself a function of the temporal resolution $\Delta t$ to be inferred.
This leads to a fundamentally self-consistent inference problem.

\subsection{Photon Statistics and Observation Noise}

We assume that during an observation window $\Delta t$, the total number of detected
photons follows Poisson statistics,
\begin{equation}
N_\gamma \sim \mathrm{Poisson}(\Phi\, \Delta t),
\end{equation}
where $\Phi$ is the photon flux (photons per unit time).

The observed spatial variance is
\begin{equation}
\sigma_{\mathrm{obs}}^2(t)
=
2Dt + \sigma_0^2,
\end{equation}
where $\sigma_0^2$ denotes the static point-spread-function (PSF) floor.

\subsection{Variance of the Variance Estimator}

For a Gaussian spatial profile estimated from $N_\gamma$ detected photons, the variance
of the maximum-likelihood estimator of the variance satisfies (see Appendix~B for a
derivation)
\begin{equation}
\mathrm{Var}(\widehat{\sigma^2})
=
\kappa \frac{\sigma_{\mathrm{obs}}^4}{N_\gamma},
\label{eq:var_sigma_photon}
\end{equation}
where $\kappa = 2$ for an ideal Gaussian PSF and asymptotically efficient estimators in
the large-photon limit. This coefficient is fixed by the statistical model and is not a
free parameter.

\subsection{Propagation of Uncertainty to Time}

Using $\sigma_{\mathrm{obs}}^2 = 2Dt + \sigma_0^2$, standard error propagation yields
\begin{equation}
\mathrm{Var}(\hat t)
=
\frac{\mathrm{Var}(\widehat{\sigma^2})}{(2D)^2}
=
\frac{\kappa}{4D^2}
\frac{(2Dt + \sigma_0^2)^2}{N_\gamma}.
\end{equation}

Substituting $N_\gamma = \Phi\, \Delta t$ gives
\begin{equation}
\mathrm{Var}(\hat t)
=
\frac{\kappa}{4D^2}
\frac{(2Dt + \sigma_0^2)^2}{\Phi\, \Delta t}.
\end{equation}

\subsection{Self-Consistent Temporal Resolution}

The minimal resolvable time interval is defined operationally by requiring that the
uncertainty of the inferred time matches the resolution being tested,
\begin{equation}
\Delta t_{\min}^2
=
\mathrm{Var}(\hat t),
\end{equation}
which is consistent with the estimator-based criterion used in previous sections up to
fixed numerical confidence factors.

This leads to the self-consistency equation
\begin{equation}
\Delta t_{\min}^3
=
\frac{\kappa}{4D^2 \Phi}
(2Dt + \sigma_0^2)^2.
\label{eq:self_consistent}
\end{equation}

\subsection{Two Asymptotic Regimes}

\paragraph{PSF-dominated regime ($\sigma_0^2 \gg 2Dt$).}

In this limit,
\begin{equation}
\Delta t_{\min}
=
\left(
\frac{\kappa\, \sigma_0^4}{4D^2 \Phi}
\right)^{1/3}.
\end{equation}

This yields the nontrivial scaling law
\[
\Delta t_{\min} \propto \Phi^{-1/3},
\]
which is fundamentally distinct from the standard $1/\sqrt{N}$ noise suppression.

\paragraph{Diffusion-dominated regime ($2Dt \gg \sigma_0^2$).}

Here,
\begin{equation}
\Delta t_{\min}
=
\left(
\frac{\kappa\, t^2}{\Phi}
\right)^{1/3}
=
\kappa^{1/3}\,
\frac{t^{2/3}}{\Phi^{1/3}}.
\end{equation}

Temporal resolution degrades algebraically with absolute time, reflecting the loss of
distinguishability as diffusion broadens.

\subsection{Interpretation}

The cubic self-consistency relation~\eqref{eq:self_consistent} is the central result of
the photon-limited regime. It expresses a closed information-theoretic loop:
\begin{quote}
\emph{Improving temporal resolution requires photons, but photons require time.}
\end{quote}

The exponent $1/3$ is therefore not heuristic but a direct consequence of Poisson photon
statistics combined with quadratic spatial uncertainty.

\subsection{Relation to Fisher Information}

The photon-limited bound can alternatively be derived from a Fisher-information
calculation in which the photon number $N_\gamma=\Phi\Delta t$ is treated as a
stochastic, time-dependent resource. Both approaches yield identical scaling and
consistent numerical prefactors under matching modeling assumptions, confirming the
internal consistency of the framework.
